Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2024-02-29 revision/review_response.qmd

Beck et al. developed a new method for estimating metabolism in estuaries using continuous dissolved oxygen data. Metabolism is a key component of ecosystems, yet it is inherently difficult to estimate, especially in estuaries due to tidal currents. Their method is based on a stream Bayesian approach, and the authors expanded it by adjusting the time interval over which model parameters are estimated. The longer intervals allow for estimation over tidal cycles, which acts as noise in the time series, and allows for improved model estimation. The authors also compared their model to a more conventional approach (i.e., Odum model) and show the model’s improved estimation, specifically avoiding impossible results, such as negative production.

The model is written in the programming language R, making it widely available to the public. The online documentation and vignettes are clear and show the users key features and examples of the package. This is a nice bonus to go along with a well-written manuscript and expand the impact of the products.

I enjoyed reading the manuscript. It is technical where it needs to be, but also provides general context for the problem and the rationale for the technique. I was tripped up during a few of the sections, and I provide some suggestions that might improve the flow of the article. I’m also wondering if this model approach is only applicable to estuaries, or if it could be useful for other aquatic ecosystems? Overall, the authors did a great job. Nice work!

*Response: Thank you for providing thoughtful comments on our manuscript.  Please see our responses below.*

@@

*Response: Yes, it is possible with caveats.  Please see our response to reviewer one on line 188.  Specifically, additional comparisons are needed and the default uninformed priors based on estuarine literature must be reconsidered.*

Model parameter b.

I’m less familiar with this term, and the authors could do a little bit more to explain what it is. On L143, they say that it is “fairly well known for open for open-ocean conditions”, but they never actually define the term. Looking at Eq 2, it seems to relate to “gas transfer efficiency” given that it is a coefficient on the wind-speed/viscosity term. That being said, I’m also a little confused on how Eq 3 is integrated into the model framework. Eq. 3 seems to relate more closely with streams as water depth is included as a scaler for kw. So somehow the two models (one more typical for lentic systems, the other more typical for lotic systems) are integrated together. In theory this makes sense in an estuary as both wind and flow likely affect gas transfer.

*Response: Apologies for this confusion.  Eq. 3 is not used in EBASE, rather it is discussed as the gas transfer used in BASE that is inappropriate for estuaries, justifying our use of the gas exchange model in @Wanninkhof14.  Line 166 was revised to make this more clear: "Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein.  Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE (Equation 2)..."*

@@

L26: which open-source software? If only one (i.e., R), then can list it here.

*Response: Added R in parentheses.*

L40: Organic matter produced in the estuary has other outcomes. Re-respired, consumed and incorporated into biota, exported to the ocean, etc.

*Response: Sentence was revised as follows: "...forming particulate organic carbon with multiple fates, e.g., buried in estuarine sediments, incorporated into biota, transferred to the ocean, etc. [@hu2006; @middelburg2007]."*

L83: which computed rates were negative? P, R, or both?

*Response: Revised as "The computed rates were negative 23% of the time for both P and R..."*

L128: mean depth?

*Response: The software can handle a single value (e.g., mean depth) or a vector of tidal height depths equal in length to the input time series. This is expressed in the documentation: https://fawda123.github.io/EBASE/reference/ebase.html.  This is also expressed in the new table showing model inputs/outputs as requested above.*

@@

L133: Thanks for providing the units. I am curious about area and time. I’m assuming that area is surface water area and time is the interval of the dissolved oxygen data. If so, perhaps expand on these so the author knows what the temporal and horizontal spatial scales are.

*Response: See the new table for inputs/outputs.*

L134: “R is constant”. I believe R is allowed to vary across days, but R is constant within a day. Please specify.

*Response: We have specified that the unknown metabolic parameters ($a$, $R$, and $b$) are constant at the unit of the model optimization period chosen by the user, although P and D vary continuously given their dependence on measured input parameters (e.g., PAR, wind) that can change at each time step.  Please see our response to reviewer 1 on line 181.*

@@

L143: what is b? and for what time interval is b constant? This ends up being an output (Figure 3) and it is confusing to say that it is constant. So maybe it is constant within a day? Is this a standard parameter for gas exchange? As with the other parameters, providing some type of ‘name’ would be helpful, especially if provided in a table.

*Response: Please see our response to the previous comment and the new input/output table.  It's an ambiguous parameter, but really just an indication of how sensitive the gas transfer is to wind speed.  The sentence was revised as "The $b$ parameter is fairly well known for open-ocean conditions primarily as an expression of the sensitivity of gas exchange to wind..."*

@@

L164: The equation implies that Kw would increase as water level (H) increases. It seems the authors have this statement inversed or talking about measured reaeration (K). I believe this relates to streams where increasing water levels relate to higher water velocities, and thus higher turbulence and higher gas transfer velocity (kw). See General Comments.

*Response: This equation is *

L189: Thanks for putting together the web page. I’m curious if the data in this manuscript are included within the package itself, so it is possible to regenerate the results entirely? I’m also wondering how stable this website is and if any online information that is described in the manuscript will change? I’m not that familiar with these online reference materials, but I’m thinking of a static version (tag) that will stay with the paper so a reader 10 years from now could access the exact same materials that go along with this paper.

*Response: It's the exception and not the norm that published software remains the same after a descriptive manuscript is published.  For this reason, we have added a version of EBASE that is appropriate for this paper and included the version number in the text.  ......* EDIT

### 2024-02-29 manu-draft.qmd
more revisions

Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to unphysical results and poorly constrained errors. Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation). EBASE applies a Bayesian framework to a simple process-based model and DO observations, allowing the estimation of critical model parameters, specifically light efficiency and respiration, as informed by a set of prior distributions. EBASE improves upon the stream-based model from which it was derived by accommodating missing DO data and allowing the user to set the time period over which parameters are estimated. We demonstrate that EBASE can recover known metabolic parameters from a synthetic time series, even in the presence of noise (e.g., due to tidal advection) and when prior distributions are uninformed. Optimization periods of 7 days and 30 days are more preferable than 1 day. A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided. EBASE is available using open-source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, river, ocean, and atmosphere. As materials are transported from land to rivers to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011]. Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006]. Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean. As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere (i.e., outgassing), the sediments (i.e., burial), and the ocean (i.e., advection).

@@

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS), a network of 30 estuaries across the U.S. Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate an assumption of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by [@caffrey2003]. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites. Note that R is often expressed as a negative value indicating consumption of dissolved oxygen. Here and throughout, we express correct values of R as positive values given the metabolic model described below.

@@

where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (@tbl-ebaseiotab).

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

>>>>

Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to unphysical results and poorly constrained errors. Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation). EBASE applies a Bayesian framework to a simple process-based model and DO observations, allowing the estimation of critical model parameters, specifically light efficiency and respiration, as informed by a set of prior distributions. EBASE improves upon the stream-based model from which it was derived by accommodating missing DO data and allowing the user to set the time period over which parameters are estimated. We demonstrate that EBASE can recover known metabolic parameters from a synthetic time series, even in the presence of noise (e.g., due to tidal advection) and when prior distributions are uninformed. Optimization periods of 7 days and 30 days are more preferable than 1 day. A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided. EBASE is available using open-source software (R) and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, river, ocean, and atmosphere. As materials are transported from land to rivers to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984]. For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011]. Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon with multiple fates, e.g., buried in estuarine sediments, incorporated into biota, transferred to the ocean, etc. [@hu2006; @middelburg2007]. Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean. As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere (i.e., outgassing), the sediments (i.e., burial), and the ocean (i.e., advection).

@@

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS), a network of 30 estuaries across the U.S. Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate an assumption of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by [@caffrey2003]. The computed rates were negative 23% of the time for both P and R, on average, and as high as 69% for individual sites. Note that R is often expressed as a negative value indicating consumption of dissolved oxygen. Here and throughout, we express correct values of R as positive values given the metabolic model described below.

@@

where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is fairly well known for open-ocean conditions primarily as an expression of the sensitivity of gas exchange to wind, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$. As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (@tbl-ebaseiotab).

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
